---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# mateR2

<!-- badges: start -->
<!-- badges: end -->

mateR2 is an MCMC-driven simulation engine for generating breeding matrices and
single-cohort pedigrees under precise demographic constraints. It treats the
mating pool of a dioecious species as a bipartite network, which preserves the
joint covariance between male and female reproductive output. It is aimed at the
skewed, "sweepstakes" reproductive systems common in wild aquatic populations
(e.g. salmon and sturgeon), and at evaluating how pedigree reconstruction
frameworks behave under them.

The pipeline runs in three stages:

1. **Metropolis-Hastings MCMC.** `generate_map_table()` searches a compressed
   state space of discrete mating-block configurations for a Mate-Pair Summary
   Table hitting your targets for parental abundance ($N_P$), operational sex
   ratio ($SR$) and mean number of mates ($MM$).
2. **Degree-preserving edge swapping.** `mp_table_to_matrix()` expands that table
   into an individual-level binary matrix, and `randomize_mating_structure()`
   rewires it to dissolve the block-diagonal structure *without altering any
   individual's mate count*.
3. **Fecundity and sampling.** `brd.mat.fitness()` converts the binary matrix
   **M** into a weighted breeding matrix **K**, and `mat.sub.sample()` draws a
   juvenile sample without replacement to emulate field sampling effort.

## Installation

```r
# install.packages("pak")
pak::pak("nicksard/mateR2")

# devtools also works, and is the way to force the vignettes to build:
# devtools::install_github("nicksard/mateR2", build_vignettes = TRUE)
```

Optional extras: `coda` for `run_mcmc_chains()`, `igraph` for
`count_network_components()`.

## A worked example

```{r stage1, message = FALSE}
library(mateR2)

# Stage 1: 100 parents, 2:1 sex ratio, 2 mates each
fit <- generate_map_table(
  Np_target = 100, sr_target = 2.0, mean_mates_target = 2.0,
  max_males_per_female = 10, max_females_per_male = 10,
  np_weight = 50, sr_weight = 50, mm_weight = 50, decay_constant = -0.05,
  n_iter = 50000, burn_in = 5000, thin = 10, seed = 42
)
unlist(fit$map_stats)
```

```{r stage2}
# Stage 2: expand to individuals, then rewire at 25% mixing intensity
M <- mp_table_to_matrix(fit$map_table)
M <- randomize_mating_structure(M, I = 0.25)
M[1:4, 1:6]
```

Rewiring dissolves the block structure while leaving every marginal untouched:

```{r topology}
mat.stats(M)[, c("num.males", "num.females", "overall.mean.mates")]
count_network_components(mp_table_to_matrix(fit$map_table))  # before
count_network_components(M)                                  # after
```

```{r stage3}
# Stage 3: allocate fecundity, then sample 500 juveniles
K <- brd.mat.fitness(M, min.fert = 1000, max.fert = 5000, type = "uniform")
ped <- convert2ped(mat.sub.sample(K, num_offspring = 500))
head(ped, 3)
```

```{r stats}
# Sibship structure and parental detection in the sample
sampled <- ped2mat(ped)
sib.stats(sampled)[, c("mean_FS", "mean_MHS", "mean_PHS")]
parent.class.stats(sampled)[, c("detected", "singletons", "doubletons")]
```

## Reproducibility

The sampler draws from R's random number generator, so `set.seed()` governs a
run, and passing `seed` to `generate_map_table()` does the same. Identical seeds
produce identical chains across platforms.

`run_mcmc_chains()` runs several dispersed chains and returns `coda` traces with
Gelman-Rubin and effective sample size diagnostics.

## Getting started

The quick-start vignette walks through the same pipeline in more detail:

```r
vignette("quickstart-breeding-sim", package = "mateR2")
```

## Citation

```r
citation("mateR2")
```
