---
title: "Choosing a Complexity Penalty"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Choosing a Complexity Penalty}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

## What the penalty is for

Many different mating block configurations satisfy the same macro-demographic
targets. A population of 400 parents at a sex ratio of 2 with a mean of 2 mates
can be assembled from many simple pair bonds, from a few dense communal groups,
or from anything in between. The MCMC sampler has no reason to prefer one over
another on demographic grounds alone, so `generate_map_table()` adds a
structural complexity penalty to the objective: a non-positive decay constant
`decay_constant` (written $\gamma$ below) multiplied by the total complexity of
the network, where each block type $b$ carries a complexity score $\delta_b$.

The intent is to express a prior over network architectures. Setting
$\gamma = 0$ removes the penalty entirely and lets the demographic targets alone
decide.

What the penalty can and cannot express turns out to be constrained by algebra,
and this vignette works through it, because the choice of metric and decay
constant is easy to get wrong in a way that is invisible in the demographic
diagnostics.

## The constraint the penalty operates under

Total parental abundance and total edge count are both pinned by the targets.
$N_P$ is the target, and the edge count follows from $N_P$ and mean mates. So
the penalty cannot change how much mating happens. It can only change how that
fixed amount of mating is distributed across blocks.

That has a consequence for which metrics can do anything at all. Consider the
default, $\delta_b = m + f - 2$. Summed over the whole network this is

```
sum over blocks of  c_b (m_b + f_b - 2)  =  N_P  -  2 * (number of blocks)
```

Since $N_P$ is fixed, minimising this quantity is exactly equivalent to
*maximising the number of blocks*. The penalty is not expressing a preference
about density at all. It is expressing a preference for many blocks, which given
a fixed $N_P$ means many small ones.

The same collapse happens for any $\delta$ linear in $m$ and $f$, and for
bilinear forms, because $\sum_b c_b m_b f_b$ is the edge count and is also
fixed. Only superlinear metrics can express anything about density. `mateR2`
therefore ships four:

| `complexity` | $\delta_b$ | what it can express |
|---|---|---|
| `"sum"` | $m + f - 2$ | block count only (the default) |
| `"asymmetry"` | $\lvert m - f \rvert$ | sex balance within a block |
| `"cyclomatic"` | $(m-1)(f-1)$ | independent cycles, so communal structure |
| `"quadratic"` | $(m + f - 2)^2$ | block size, superlinearly |

## Measuring what actually happens

The quantity of interest is the densest block the sampler settles on, measured
as the number of pairings it contains ($m \times f$), together with the share of
blocks that are communal in the sense that both sexes mated more than once.

```{r sweep}
library(mateR2)

probe <- function(gamma, metric, seed = 1) {
  m <- generate_map_table(
    Np_target = 200, sr_target = 2, mean_mates_target = 2,
    max_males_per_female = 10, max_females_per_male = 10,
    decay_constant = gamma, complexity = metric,
    np_weight = 200, sr_weight = 200, mm_weight = 200,
    n_iter = 1e5, burn_in = 1e4, thin = 100,
    seed = seed, verbose = FALSE, show_progress = FALSE
  )
  tb       <- m$map_table[m$map_table$MAP_Count > 0, ]
  edges    <- tb$Males * tb$Females
  communal <- tb$Males > 1 & tb$Females > 1
  data.frame(
    gamma, metric,
    densest_block = max(edges),
    largest_block = max(tb$Males + tb$Females),
    pc_communal   = round(100 * sum(tb$MAP_Count[communal]) / sum(tb$MAP_Count)),
    MM_error_pc   = round(100 * abs(m$map_stats$MeanMates - 2) / 2, 1)
  )
}

grid <- expand.grid(
  gamma  = c(0, -0.05, -0.25),
  metric = c("sum", "asymmetry", "cyclomatic", "quadratic"),
  stringsAsFactors = FALSE
)
res <- do.call(rbind, Map(probe, grid$gamma, grid$metric))
res[order(res$metric, -res$gamma), ]
```

Read the `densest_block` column down each metric. For `sum`, `asymmetry` and
`cyclomatic` it does not move at all as $\gamma$ strengthens. Whatever the
penalty is doing to the rest of the network, the densest block is untouched.

This is the algebra above showing up in practice. Strengthening the penalty
makes most blocks simpler, and because the total amount of mating is fixed, the
multi-mating it displaces has to go somewhere. It concentrates into the few
blocks that remain, so the network becomes more sharply divided between many
pair bonds and a few dense clusters rather than uniformly simpler. The
concentration is easier to see at larger $N_P$: repeating this sweep at 400
parents drops the communal share from 5% to between 1% and 3% while
`densest_block` holds at 48 to 50 throughout.

The practical consequence is that a strengthening `sum` penalty should not be
read as a control on how dense the densest mating group is allowed to be. It
is not one.

`quadratic` is the exception, and it is the only metric here that penalises
block size superlinearly. It does reduce the densest block, which is what makes
it worth having.

## Choosing a decay constant

The usable range of $\gamma$ is narrower than it looks, and it interacts with
scale.

```{r gamma-range}
strong <- do.call(rbind, lapply(c(-0.25, -0.5, -1), probe, metric = "quadratic"))
strong
```

Beyond a certain strength the penalty stops being a prior over architectures and
starts overriding the demographic targets. Watch `MM_error_pc`: once it rises,
the sampler is buying structural simplicity by missing the mating system it was
asked to produce, which is almost never the trade a user wants.

Where that boundary falls depends on $N_P$. At the 200-parent scale used above,
`quadratic` at $\gamma = -0.25$ suppresses density while holding mean mates. At
400 parents the same setting drove the mean-mates error above 20%. Any
$\gamma$ chosen from a small pilot should be re-checked at the scale you intend
to run.

For this reason `"sum"` at $\gamma = -0.05$ remains the package default. It is a
mild preference for simpler blocks that does not perturb the demographic
targets at any scale tested. It should be understood for what the algebra says
it is — a preference for more blocks — and not as a control on density.

## Practical guidance

Use $\gamma = 0$ when the demographic targets are the whole specification and
network architecture is not of interest. The sampler will find whatever
configuration fits.

Use `"sum"` at a mild $\gamma$ around $-0.05$ when simple pair bonds should be
the default architecture and occasional dense clusters are acceptable. This is
the default and suits most single-cohort simulations.

Use `"cyclomatic"` when the biological concern is communal mating specifically
rather than block size. It scores $(m-1)(f-1)$, which is zero for any block
where one sex mated once, so it leaves one-to-many fans alone and charges only
for configurations where both sexes multiply mate.

Use `"quadratic"` when there is a hard biological ceiling on how many
individuals can plausibly interact in one mating group. Check the mean-mates
error at your own $N_P$ before committing, because this is the metric most
likely to start overriding the targets.

Whichever you choose, report both the metric and the decay constant alongside
the demographic targets. Two runs with identical $N_P$, sex ratio and mean mates
can have visibly different family structure, and the sibship distributions that
downstream reconstruction software sees will differ accordingly.
