---
title: "Choosing Mate Caps"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Choosing Mate Caps}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

## What the caps actually control

`generate_map_table()` and `run_mcmc_chains()` take two caps,
`max_males_per_female` and `max_females_per_male`. They look like biological
constraints, and they are, but they also do something less obvious: together
they define the entire state space the sampler searches. A block type is a pair
$(m, f)$ with $m$ up to the first cap and $f$ up to the second, so the number of
block types is simply their product. Raising a cap does not merely permit denser
mating groups, it enlarges the space through which a single-step random walk has
to travel.

That gives the caps a dual role, and the two roles pull in different directions.

## The caps are bounded below by the targets

The caps cannot be set freely. In a block where every male mates with every
female, each male has $f$ mates and each female has $m$ mates, so the caps must
be at least the mean mating success the demographic targets imply:

```
mean mates of a male    =  2 * MM / (SR + 1)
mean mates of a female  =  2 * MM * SR / (SR + 1)
```

`check_target_viability()` enforces the related graph-theoretic floor, refusing
targets where a male would need fewer than one mate. What it does not do is tell
you how much headroom to leave above the requirement, and that is the choice
this vignette is about.

It is worth being explicit that a flat cap is *not* a constant amount of
headroom. A cap of 10 against a requirement of 1.0 mate is ten times the
requirement; the same cap against a requirement of 5.3 is under twice. A single
number applied across a demographic grid therefore imposes a quite different
search problem in every cell.

## Measuring the trade

Define $k$ as the cap expressed as a multiple of the requirement, and sweep it
across six mating systems spanning the feasible region.

```{r sweep}
library(mateR2)

NP <- 200; W <- NP / 2      # linear weight rule

probe <- function(SR, MM, k, seed = 1) {
  req_male   <- 2 * MM / (SR + 1)
  req_female <- 2 * MM * SR / (SR + 1)
  cap_f_per_m <- max(1, ceiling(k * req_male))
  cap_m_per_f <- max(1, ceiling(k * req_female))
  r <- run_mcmc_chains(
    Np_target = NP, sr_target = SR, mean_mates_target = MM,
    max_males_per_female = cap_m_per_f, max_females_per_male = cap_f_per_m,
    n_chains = 4, n_iter = 1e5, burn_in = 1e4, thin = 100, seed = seed,
    np_weight = W, sr_weight = W, mm_weight = W, show_progress = FALSE
  )
  d <- r$diagnostics
  data.frame(SR, MM, k,
    caps       = paste0(cap_m_per_f, "/", cap_f_per_m),
    block_types = cap_m_per_f * cap_f_per_m,
    Np_err     = round(d$error_pct[d$parameter == "Np"], 2),
    MM_err     = round(d$error_pct[d$parameter == "mean_mates"], 2),
    max_rhat   = round(max(d$rhat), 3),
    min_ess    = round(min(d$ess)))
}

sc   <- data.frame(SR = c(1, 1, 1, 2, 2, 4), MM = c(1, 2, 4, 2, 4, 4))
grid <- expand.grid(k = c(1, 1.5, 2, 4), row = 1:6)
out  <- do.call(rbind, Map(function(k, i) probe(sc$SR[i], sc$MM[i], k),
                           grid$k, grid$row))
out[order(out$SR, out$MM, out$k), ]
```

Averaged across the six systems:

```{r summary}
aggregate(cbind(Np_err, MM_err, min_ess, max_rhat) ~ k, out, mean)
```

Two things to read off this.

**Accuracy improves with headroom, and keeps improving.** Both the parental
abundance error and the mean mates error fall as $k$ rises, and mean mates is
much the more sensitive of the two. At $k = 1$ the sampler is confined to a
space barely large enough to contain a valid answer, and it misses the mating
system badly in the balanced-sex-ratio cells. There is no accuracy argument for
a tight cap.

**Mixing does not improve with headroom. It peaks and then falls.** Effective
sample size rises from $k = 1$ to $k = 2$ and then drops sharply at $k = 4$,
while the Gelman-Rubin statistic degrades. This is the cost that pays for the
accuracy gain, and it is the reason the caps cannot simply be set generously and
forgotten.

## The mixing cost is an iteration budget, not a bad cap

It would be easy to conclude that large caps are harmful. They are not. The
degradation at $k = 4$ is a consequence of holding the iteration count fixed
while the state space grows with the product of the caps. Take the worst cell
above and vary only the run length:

```{r budget}
budget <- function(n) {
  r <- run_mcmc_chains(
    Np_target = 200, sr_target = 2, mean_mates_target = 4,
    max_males_per_female = 22, max_females_per_male = 11,
    n_chains = 4, n_iter = n, burn_in = n / 10, thin = 100, seed = 1,
    np_weight = 100, sr_weight = 100, mm_weight = 100, show_progress = FALSE
  )
  d <- r$diagnostics
  data.frame(n_iter = n, max_rhat = round(max(d$rhat), 3),
             min_ess = round(min(d$ess)))
}
do.call(rbind, lapply(c(1e5, 5e5, 2e6), budget))
```

The same caps that fail to converge on a short run converge cleanly on a longer
one. The cap is not the problem; the ratio of iterations to state-space size is.
So the caps and the run length are a single decision, not two independent ones.

## A degenerate case worth recognising

In the sweep above, the row with `SR = 1, MM = 1, k = 1` gives caps of 1 and 1,
a state space of a single block type, an $\hat{R}$ of `NaN` and an effective
sample size of zero. That is not a sampling failure. With only 1:1 blocks
available there is exactly one configuration satisfying the targets, every chain
finds it, and the within-chain variance is identically zero, which is what makes
the diagnostics undefined. The demographic errors on that row are the lowest in
the table.

This matters for interpreting diagnostics generally. A chain pinned against a
structural boundary can look unconverged by the usual statistics while having
found the exact answer. Where $\hat{R}$ is undefined or extreme, check the
realized demography and the number of distinct states visited before concluding
the sampler misbehaved.

## Practical guidance

Set the caps as a multiple of the requirement implied by your targets, not as a
flat number, unless the flat number is genuinely a biological ceiling. A
multiple of roughly 1.5 to 2 is the efficient choice at a fixed iteration
budget: it captures nearly all of the accuracy available and sits at or near the
peak of the mixing curve.

Go higher when the biology demands it, for instance when a species is known to
form large communal groups that a tighter cap would exclude by construction.
When you do, raise the iteration count with it. The state space grows with the
product of the caps, so doubling both quadruples the space, and a run length
that was ample before will not be.

Report the caps alongside the demographic targets. Two runs with identical
$N_P$, sex ratio and mean mates but different caps are searching different
spaces and can produce visibly different block structure.

Finally, always read $\hat{R}$ and effective sample size together with the
realized demography. The cells where the diagnostics look worst are sometimes
the cells where the answer is exact.
